---
title: "Getting Started with filtertools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with filtertools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(filtertools)
library(dplyr)
```

## Why filtertools?
When exploring data interactively, you often need to filter rows based on string patterns or NA values. With base dplyr, this can get verbose:

```{r verbose-dplyr, eval=FALSE}
# Keep rows where name contains "John"
data %>% filter(str_detect(name, regex("John", ignore_case = TRUE)))

# Remove rows where status contains "inactive"
data %>% filter(!str_detect(status, regex("inactive", ignore_case = TRUE)))

# Remove rows where all of these columns are NA
data %>% filter(!if_all(c(age, income), is.na))
```

filtertools provides intuitive wrappers that make these operations more readable:

```{r filtertools-examples, eval=FALSE}
data %>% filter_in(name, "John")
data %>% filter_out(status, "inactive")
data %>% filter_out_na(age, income)
```

## String Pattern Filtering

### filter_in(): Keep matching rows

Use `filter_in()` to keep rows where a column matches a pattern:

```{r filter-in-basic}
fruits <- tibble(
 name = c("apple", "banana", "cherry", "apricot", "blueberry"),
 color = c("red", "yellow", "red", "orange", "blue")
)

# Keep fruits containing "berry"
fruits %>% filter_in(name, "berry")

# Keep fruits starting with "a"
fruits %>% filter_in(name, "^a")
```

### filter_out(): Remove matching rows

Use `filter_out()` to remove rows where a column matches a pattern:

```{r filter-out-basic}
# Remove fruits containing "apple"
fruits %>% filter_out(name, "apple")

# Remove red fruits
fruits %>% filter_out(color, "red")
```

### Case sensitivity

By default, matching is case-insensitive:
```{r case-insensitive}
fruits %>% filter_in(name, "APPLE")
```

Set `ignore_case = FALSE` for case-sensitive matching:

```{r case-sensitive}
fruits %>% filter_in(name, "APPLE", ignore_case = FALSE)
```

### Regex patterns

The `pattern` argument accepts full regular expressions:

```{r regex-patterns}
# Starts with "a" or "b"
fruits %>% filter_in(name, "^[ab]")

# Ends with "y"
fruits %>% filter_in(name, "y$")

# Contains "rr" or "pp"
fruits %>% filter_in(name, "rr|pp")
```

## NA Handling in String Filters

By default, `filter_in()` and `filter_out()` **preserve rows with NA values**. This prevents accidental data loss during exploratory analysis:

```{r na-handling-default}
data_with_na <- tibble(
 status = c("active", "inactive", "pending", NA, "active"),
 value = 1:5
)

# NA rows are kept by default
data_with_na %>% filter_out(status, "inactive")
```

Use `na_rm = TRUE` to exclude NA rows:

```{r na-handling-remove}
data_with_na %>% filter_out(status, "inactive", na_rm = TRUE)
```

## NA Filtering

### filter_out_na(): Remove rows with NA

Remove rows where specified columns contain NA:

```{r filter-out-na}
messy_data <- tibble(
 a = c("x", "y", NA, NA),
 b = c("x", NA, "z", NA),
 c = c(1, 2, 3, 4)
)

messy_data
```

By default (`if_any_or_all = "if_all"`), rows are removed only when **all** specified columns are NA:

```{r filter-out-na-all}
# Remove rows where BOTH a and b are NA
messy_data %>% filter_out_na(a, b)
```

Use `if_any_or_all = "if_any"` to remove rows where **any** specified column is NA:

```{r filter-out-na-any}
# Remove rows where a OR b is NA
messy_data %>% filter_out_na(a, b, if_any_or_all = "if_any")
```

### filter_in_na(): Keep rows with NA

The inverse operation - keep only rows containing NA:

```{r filter-in-na}
# Find rows where any column has NA
messy_data %>% filter_in_na(a, b, if_any_or_all = "if_any")

# Find rows where all specified columns are NA
messy_data %>% filter_in_na(a, b, if_any_or_all = "if_all")
```

### Using tidyselect

Both functions support tidyselect syntax:

```{r tidyselect}
# Check all columns
messy_data %>% filter_out_na(everything(), if_any_or_all = "if_any")

# Check columns starting with a pattern
messy_data %>% filter_in_na(starts_with("a"), if_any_or_all = "if_any")
```

## Finding Non-Numeric Values

`filter_out_numeric()` helps identify rows with non-numeric values - useful for data cleaning:

```{r filter-out-numeric}
survey_data <- tibble(
 age = c("25", "thirty", "42", NA, "38"),
 score = c("100", "85", "N/A", "92", "88")
)

# Find rows with any non-numeric values
survey_data %>% filter_out_numeric()

# Check only the age column
survey_data %>% filter_out_numeric(.cols = age)

# Exclude NA from results
survey_data %>% filter_out_numeric(.cols = age, na_rm = TRUE)
```

## Filtering Character Vectors

`str_filter()` works directly on character vectors (not data frames):

```{r str-filter}
fruits <- c("apple", "banana", "cherry", "apricot")

# Keep items containing "a"
str_filter(fruits, "a")

# Keep items NOT starting with "a"
str_filter(fruits, "^a", negate = TRUE)
```

## Summary

| Function | Purpose | NA behavior |
|----------|---------|-------------|
| `filter_in()` | Keep rows matching pattern | Preserves NA by default |
| `filter_out()` | Remove rows matching pattern | Preserves NA by default |
| `filter_in_na()` | Keep rows with NA | - |
| `filter_out_na()` | Remove rows with NA | - |
| `filter_out_numeric()` | Keep non-numeric rows | Includes NA by default |
| `str_filter()` | Filter character vectors | - |
